The Evolution of Phenotypically Invariant Genetic Networks

### 2017-08-02 kalman_walk_paper.tex
first pass through with comments

\section*{Evolutionary conservation of phenotype does not entail conservation of the underlying molecular mechanism leading to the rapid formation of new species}
{\centering
Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
$^{\dagger}$University of Southern California, Los Angeles, California \qquad $^{\ddagger}$University of Oregon, Eugene, Oregon \\
\texttt{jsschiff@usc.edu} \qquad \texttt{plr@uoregon.edu}
\\
}

@@

\begin{abstract}
We introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks. The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes. Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set. We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit. We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
\end{abstract}

@@

@@

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}. This is especially pertinent given the staggering complexity and scope of contemporary research programs. This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice, as models allow the development of concrete numerical predictions. Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

@@

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations'', or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Although this may frustrate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

  In this paper we outline a theoretical framework to study the evolution of biological systems. We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), however presently, we focus solely on neutral evolution, that is where phenotype is conserved over evolutionary time.

  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

@@

Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. We define an organism's phenotype $\phi(t)$ to be the molecular time-dynamics directly relevant to survival. These dynamics are the result of the interconnections of a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) and an environmental input $u(t)$.

@@

  The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment. $C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.

@@

@@

  where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system. As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.

@@

    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
    \end{align*}
    where $\dot{\kappa}$ denotes the time derivative. The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time. If we record the regulatory coefficients in the matrix
    \begin{align*}
      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
    \end{align*}
    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
      \begin{align*}
      \dot{\kappa} &= A \kappa(t) + B u(t) .
      \end{align*}
      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
      \begin{align*}
      \phi(t) &= C \kappa (t)
      \end{align*}
      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. (Note: if the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

      Since the input is simply an impulse, its phenotype is equivalent to its impulse response

      \begin{align*}
        \phi(t) = h(t) = \sin(t) + \cos(t)  .
      \end{align*}

@@

  \section*{Linear Evolutionary Systems}

  The literature is filled with detailed observations of molecular systems and their diversity. There are examples of significant diversity in the networks underlying processes such as circadian rhythm \citep{sancar2008intelligent}, cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, pattern formation, and metabolism  \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level. How many different mechanisms have the same function? We urge the reader to consider the metaphorical black box.

  Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics). Any linear and minimal system (a gene network) -- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes -- has an identical phenotype up to a change of coordinates. 
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= \bar{C} e^{\bar{A} t} \bar{B}
  \end{align}
  Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, have the same phenotype if they are related by a change of coordinates.

  Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
  \begin{align}
    VB &= B \implies \bar{B} = B \\
    CV &= C \implies \bar{C} = C
  \end{align}
  In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.

  Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
  and a \emph{Linear Evolutionary System} is, 
  \begin{align}
    \mathcal{S}(\tau) := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
  That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.

  More generally, \plr{introduce Kalman},
  we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
  \begin{equation} \label{eqn:equivalence}
      \begin{aligned}
      \calA_n(A_0) 
          &= \left\{
              A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
          \right\}  \\
          &= \left\{
              A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
          \right\} .
      \end{aligned}
  \end{equation}
  Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
  Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

  Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$. This set can be completely parameterized using the \emph{Kalman Decomposition}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined. Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.

@@

    Let
    \begin{align*}
      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
    \end{align*}
    with $V(\tau)$ preserving both $B$ and $C$, then
    the set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in Example \ref{ex:oscillator} are given by
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

\end{tikzpicture}
    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value}
    \end{figure}
 % \end{example}

@@

@@

@@

  At any given time, there will be a range of network coefficients present in the population
  due to segregating genetic polymorphism.
  Over many generations, even if selective pressures do not change,
  this range of networks will shift 
  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
  How much variation do we expect to find within a population?
  Is this range limited by available variation or kept in check by selection?
  How fast will a population explore the space of equivalent networks?
  To answer these questions, we need to know two things known at best only roughly \citep{felsenstein1988phylogenies} --
  the strength of stabilizing selection on the phenotype,
  and the amount (and structure) of heritable variation in the genotype.
  We aim to get order-of-magnitude estimates.

>>>>

\section*{Gene regulatory network drift and speciation occurs rapidly under neutrality}
{\centering
Joshua S. Schiffman$^{\dagger}$ \qquad Peter L. Ralph$^{\dagger \ddagger}$ \\
$^{\dagger}$University of Southern California, Los Angeles, California \qquad 
$^{\ddagger}$University of Oregon, Eugene, Oregon \\
\texttt{jsschiff@usc.edu} \qquad 
\texttt{plr@uoregon.edu}
\\
}

% Other title ideas:
% More than one way to grow a cat: regulatory network redundancy and speciation
% 
% Neutral network drift leads to incompatibilities on the same time scale as genetic drift
% 
% More than one way to grow a cat: can network drift lead to speciation?
% 
% How fast does network drift create incompatibilities?
% 
% Beyond the snowball: a quantitative model of incompatibility accumulation
% 
% Systems drift and speciation: more than one way to grow a cat
% 
% An explicit model of neutral regulatory network evolution, with applications
% to the rate of accumulation of hybrid incompatibility
%

@@

\begin{abstract}
Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks. The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes. Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set. We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit. We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
\end{abstract}

@@

\plr{Note: in the \LaTeX source I'm putting in semantic linebreaks, so it's easy to edit and move around phrases and ideas.}

@@

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. 
Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. 
\plr{Saying "stalled" reflects negatively on other fields - rephrase?}
There does exist a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1, weiss2000phenogenetic, edelman2001degeneracy}. 
\plr{I don't think the HW example will speak to our audience.  Also, better to say why math is good rather than why not-math is bad.}
However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze, servedio2014not}. 
This is especially pertinent given the staggering complexity and scope of contemporary research programs. 
This outlook necessitates the advancement of conceptual frameworks of such precision, 
only mathematics will suffice, as models allow the development of concrete numerical predictions. 
Previously it has been suggested that any idealized study of evolution is incomplete 
without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}. 
\plr{probably some more recent Wagner papers in this line as well?}
%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

@@

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations \citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary, draghi2015robustness} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative, chertkova2017insilico}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. 
\plr{not clear what you mean by "actually"}
Models with different parameter schemes, or even different mechanics can be equally accurate, 
but still not \emph{actually} agree with what is being modelled. 
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
In chemical engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. 
In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.  
\plr{here you are jumping to evolution -- haven't said yet why evolution "can explore" these -- needs to say explicitly what we mean by neutral here}
Although this may frustrate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; 
macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level 
\citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

In this paper we outline a theoretical framework to study the evolution of biological systems. 
\plr{rather than "we expect" maybe say that the framework could be applied to non-neutral evolution}
We expect wide applicability to this approach (\emph{i.e.} non-neutral evolution), 
however presently, we focus solely on neutral evolution, 
\plr{here's the definition of neutral, but it's not quite right?}
that is where phenotype is conserved over evolutionary time.

\plr{We don't "derive" it, we point out that Kalman already derived it.}
We derive an analytical description of the set of all linear biological systems with identical phenotypes 
\plr{We need something here saying what we mean by "gene network architecture".  Also, I don't think we claim anything for "all biological systems".}
-- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. 
In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, 
and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. 
Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

\plr{Below we mix general language with language specific to GRNs, like switching between "internal state" and "transcription factor concentration.  
Rather than keep making comments like "(or some other system)",
we could say up top that this applies to other situations, but to make it concrete we'll talk about regulatory networks.}

@@

Here we outline a method to model biological systems, such as gene regulatory networks, as linear dynamical systems. 
We define an organism's ``phenotype'' to be the time dynamics of molecular concentrations directly relevant to survival, 
\plr{tried to say it more precisely. better phrase than "component of phenotype"?}
and denote by $\phi_k(t)$ the concentration of the $k^\text{th}$ component of phenotype at time $t$.
These dynamics are the result of the interconnections of 
a gene regulatory network (or some other biological system) $A$ (an $n \times n$ matrix) 
and an environmental input $u(t)$.
\plr{say what $u$ is (a vector)}
\plr{say in words what $\kappa$ is before the equation.

@@

The change in internal molecular concentrations as a function of time $\dot{\kappa}(t)$ is simply a function of the systems current state (or kryptotype), the system architecture $A$ -- such as a gene regulatory network, the environmental input $u(t)$, and $B$ (an $n \times l$ matrix) -- how the system processes its environment. 
\plr{Could say instead of "filters" that e.g. $C_{ij}$ is the amount by which the $j$th input $u_j$ affects the production of the $i$th transcription factor concentration.}
$C$ (an $l \times n$ matrix) filters the dynamics relevant to survival.
\plr{This "Thus" needs expanding, e.g., "the solution to this equation is unique and given by" with a reference.}

@@

\plr{You've missed out $\kappa(0)$. Also, you can't have $t$ be both a limit of the integral and the variable being integrated over -- check this and get it right.}

@@

where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system.

\plr{How about expanding this minor comment into a short paragraph with more interpretation of not just $A$ but also $B$ and $C$? Should say why. 
Include talking through why $A_{ij}>0$ means that $j$ "upregulates" $i$.}
As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.

@@

\plr{you haven't explained $u(t)$ above. and, isn't it more natural in this set-up to have $B = [1,0]$ than $[1,1]$?}
    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
    \end{align*}
    where $\dot{\kappa}$ denotes the time derivative. The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time. If we record the regulatory coefficients in the matrix
    \begin{align*}
      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
    \end{align*}
    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
      \begin{align*}
      \dot{\kappa} &= A \kappa(t) + B u(t) .
      \end{align*}
      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
      \begin{align*}
      \phi(t) &= C \kappa (t)
      \end{align*}
      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. 
      (If the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

      Since the input is simply an impulse, its phenotype is equivalent to its impulse response
      \begin{align*}
        \phi(t) = h(t) = \sin(t) + \cos(t)  .
      \end{align*}

@@

  \section*{Linear Evolutionary Systems}
\plr{instead title this section "Systems drift for linear systems"?}

\plr{This paragraph feels like the introduction.}
The literature is filled with detailed observations of molecular systems and their diversity. 
\plr{Diversity doesn't imply systems drift -- only if the diverse systems are homologous.}
There are examples of significant diversity in the networks underlying processes such as 
circadian rhythm \citep{sancar2008intelligent}, 
cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, 
pattern formation \plr{cite?}, 
and metabolism \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. 
Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level. 
How many different mechanisms have the same function? 
We urge the reader to consider the metaphorical black box.

Gene regulatory networks with identical phenotypes (external dynamics) do not necessarily have identical kryptotypes (internal dynamics). 
Any linear and minimal system (a gene network) 
-- minimal, informally meaning that the system's phenotype is achieved with the fewest possible number of genes 
-- has an identical phenotype up to a change of coordinates. 
\plr{"has an identical phenotype" is not what you mean to say here}
  \begin{align}
    h(t) &= C e^{A t} B \\
    &= C V^{-1} V e^{A t} V^{-1} V B \\
    &= C V^{-1} e^{V A V^{-1} t} V B \\
    &= \bar{C} e^{\bar{A} t} \bar{B}
  \end{align}
\plr{what are we calling these -- not "biological system" -- that's too broad}
Two biological systems, $\mathcal{S} = \{ A, B, C \}$, and $\bar{\mathcal{S}} = \{\bar{A} = VAV^{-1}, \bar{B} = VB, \bar{C} = CV^{-1} \}$, 
have the same phenotype if they are related by a change of coordinates. 
\plr{is this the place to say "and only if, in the minimal dimension"?}

Although systems may not be identifiable beyond a change of coordinates, at present we are primarily interested in a subset of these systems. 
That is, systems that not only have equivalent external dynamics, but also equivalent input and output relationships. 
\plr{clarify that you mean actually having the same $B$ and $C$. (at first I read "input and output relationship" as the mapping $u \to \phi$),
    i.e. behaviour across many possible inputs.}
Formally, this means systems related by a change of coordinates (any invertible matrix $V$) that leaves $B$ and $C$ invariant:
  \begin{align}
    VB &= B \implies \bar{B} = B \\
    CV &= C \implies \bar{C} = C
  \end{align}
\plr{this sentence lacks an object (or some sentence part?)}
In other words systems with varying gene regulatory network architectures yet identical selection pressures, environment, and phenotype.

\plr{I don't think it makes sense to write things in general as a function of $\tau$,
    since in general the set of "free parameters" is complicated and we don't actually specify it.
    Better, maybe define $\mathcal{V}$ to be the set of all $V$ that preserve $B$ and $C$, then define $A(V)$ instead of $A(\tau)$...}
Define $V(\tau)$ as the parameterized change of coordinates matrix that preserves $B$ and $C$, with $\tau$ a vector of free parameters. The set of \emph{all} phenotypically invariant (minimal) gene networks is, 
  \begin{align}
    A(\tau) = V(\tau) A(0) V^{-1}(\tau) ,
  \end{align}
and a \emph{Linear Evolutionary System} is, 
\plr{why is this "evolutionary"? It is the thing that we're thinking about evolution of, but that doesn't make it evolutionary.}
  \begin{align}
    \mathcal{S}(\tau) := \left\{ \begin{array}{ll} \dot{\kappa}(t) &= A(\tau) \kappa(t) + B u(t) \\ \phi(t) &= C \kappa(t) \end{array} \right .
  \end{align}
\plr{again, lacking an object}
That is, all (linear and minimal) mechanisms capable of producing the same phenotype can be realized by a unique choice of $\tau$ in $\mathcal{S}(\tau)$.

More generally, \plr{introduce Kalman},
we denote by $\calA_n(A_0)$ the set of all $n$-dimensional systems equivalent to $A_0$:
\begin{equation} \label{eqn:equivalence}
  \begin{aligned}
  \calA_n(A_0) 
      &= \left\{
          A : C e^{At} B = C e^{A_0 t} B \; \text{for}\; t \ge 0 
      \right\}  \\
      &= \left\{
          A : C A^k B = C A_0^k B \; \text{for}\; 1 \le k \le n-1 
      \right\} .
  \end{aligned}
\end{equation}
Equivalence of the two characterizations follows from the Cayley-Hamilton theorem.
\plr{To make sense of this we need to say how $C$ and $B$ change with $n$.}
Usually, the dimension $n$ and the reference system $A_0$ is implicit and we write only $\calA$.

\plr{should say more about this decomposition here.}
Regardless of minimality, two systems, even in different dimensions, can have identical external dynamics if they are in $\mathcal{A}$. This set can be completely parameterized using the \emph{Kalman Decomposition}; the set of all linear gene regulatory networks with equivalent phenotypes can be precisely defined. Note that this implies that there is always more than one possible gene regulatory network architecture per phenotype, however we can analytically describe the complete set.

@@

\plr{this is same as above, so reference instead of writing out again?}
    Let
    \begin{align*}
      A(0) = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix}, \qquad B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix},
    \end{align*}
    with $V(\tau)$ preserving both $B$ and $C$, then
    the set of all two-gene regulatory networks phenotypically equivalent to the cell cycle control network in Example \ref{ex:oscillator} are given by
    \begin{align*}
      A(\tau) &= \begin{bmatrix} 1 & 0 \\ \tau & 1-\tau \end{bmatrix} \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ \frac{\tau}{\tau-1} & \frac{-1}{\tau-1} \end{bmatrix}  \qquad \forall \ \tau \neq 1 %\\
     %   B &= \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix} \\
       %   h(t) &= \sin(t) + \cos(t) \qquad \forall \ \tau \neq 1
    \end{align*}
    \begin{figure}[H]
    \centering
% Nice tikz!!!
\begin{tikzpicture}
\begin{scope}[every node/.style={circle,thick,draw}]
  \node (A) at (0,0) {$\kappa_{1}$};
    \node[dashed] (B) at (4,0) {$\kappa_{2}$};
    \node[shape=rectangle] (U) at (2,2) {input};
    \node[shape=rectangle] (y) at (2,-2) {output};
\end{scope}

@@

\end{tikzpicture}
    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value \plr{sentence fragment}}
    \end{figure}
 % \end{example}

@@

\plr{I think we should first put the section about quantitative drift in, and then the following section that tries to get at real values of parameters.}

@@

\plr{Below we need to separate variation in $A$ from variation in phenotype.}

@@

  At any given time, there will be a range of regulatory coefficients present in the population
  due to segregating genetic variants.
  Over many generations, even if selective pressures do not change,
  this range of networks will shift 
  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
  How much variation do we expect to find within a population?
  Is this range limited by available variation or kept in check by selection?
  How fast will a population explore the space of equivalent networks?
  To answer these questions, we need to know 
  the strength of stabilizing selection on the phenotype,
  and the amount (and structure) of heritable variation in the genotype.
  These are known at best only roughly \citep{felsenstein1988phylogenies},
  so we aim for order-of-magnitude estimates.
